---
title: "p8105_hw2_yy3439"
output: github_document
date: "2023-10-03"
---

```{r}
library(tidyverse)
library(dplyr)
```

## Problem 1

Firstly, import the `pols-month.csv` file

```{r}
pols_data = read_csv(file = "./fivethirtyeight_datasets/pols-month.csv") #import the pols-month csv file

```

Firstly,



```{r}
pols_data |>
   janitor::clean_names() |> #clean-up names to r suitable names
   separate(mon, into = c("year", "month", "day"), sep = "-") |> #separate the mon variable by "-"
   mutate(
     month = as.numeric(month), #change the char variable month to numeric
     month = month.name[month]) |> #translate numeric month number to full month names
  select(-prez_dem, -prez_gop) |> #remove the prez_dem and prez_gop variables
   pivot_longer(
    gov_gop:rep_dem, #select the remaining columns
     values_to = "president") |> #and assign the values of these variables to a new variable "president"
      select(-day) #remove the day variable
    
  
```


Secondly, 


```{r}
snp_data = read_csv(file = "./fivethirtyeight_datasets/snp.csv") #load the snp.csv file
```

### Clean the snp data


```{r}
snp_data|>
   janitor::clean_names() |> #clean-up names to r suitable names
  mutate(date = as.Date(date, format = "%m/%d/%y"))|>
   separate(date, into = c("year", "month"), sep = "-") |>#separate the date variable into year, month, and day
   mutate(
     month = as.numeric(month), #change the char variable month to numeric
     month = month.name[month])|>
  arrange(year, month)|>  #rearrange by the order of year and month
  relocate(year, month, everything())
 
```


Thirdly,


```{r}
unemploy_data = read_csv(file = "./fivethirtyeight_datasets/unemployment.csv") #load the snp.csv file
```

### Merging Datasets


```{r}
unemploy_data|>
   janitor::clean_names() |>
  pivot_longer(
    jan:dec,
    names_to = "month"
  )|>
  arrange(year, month)
  
```


## Problem 2


```{r}
library(readxl)
mr_tw_data = read_excel("./fivethirtyeight_datasets/ts.xlsx", sheet = "Mr. Trash Wheel", skip = 1) #load the excel file with Mr. Trash Wheel sheet and omit the first column with figure 
```


#### Tidy Data

```{r}
mr_trash_wheel_data = mr_tw_data |> 
  janitor::clean_names() |> #clean names to appropriate r format
  select(dumpster:homes_powered)|> #omit non-data entries 
  drop_na(dumpster) |> #omit rows with no dumpster-specific data 
  mutate(
    homes_powered = ((weight_tons*500)/30), #add another variable based on the calculation in the notes
    tw_name = "Mr. Trash Wheel", #to keep track of the records after combining datasets 
    year = as.numeric(year) #change the variable type to merge datasets
    
  )

```


Similarly, we follow the same cleanning process to organize the data for Professor Trash Wheel and Gwynnda


```{r}
professor_tw_data = read_excel("./fivethirtyeight_datasets/ts.xlsx", sheet = "Professor Trash Wheel", skip = 1)|>
   janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    homes_powered = ((weight_tons*500)/30),
    tw_name = "Professor Trash Wheel"

  )

gwynnda_tw_data = read_excel("./fivethirtyeight_datasets/ts.xlsx", sheet = "Gwynnda Trash Wheel", skip = 1)|>
   janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    homes_powered = ((weight_tons*500)/30),
     tw_name = "Gwynnda Trash Wheel"
  )
```

#### Combine Datasets

Finally, combining the above three datasets

```{r}
trash_wheel = bind_rows(mr_trash_wheel_data, professor_tw_data, gwynnda_tw_data) |>
  select(trash_wheel_name = "tw_name", everything())
```


#### Discussion

The `Mr Trash Wheel` sheet has `r nrow(mr_trash_wheel_data)` observations. The `Professor Trash Wheel` sheet has `r nrow(professor_tw_data)` observations, and the `Gwynnda Trash Wheel` sheet has `r nrow(gwynnda_tw_data)` observations. Specifically, all these data is calculated after we omit rows with no specific dumpster data. 
In addition, the final combined dataset `trash_wheel` contains `r nrow(trash_wheel)` observations and `r ncol(trash_wheel)` columns. 
Each sheet contains similar key variables specifying various types of dumpsters: plastic bottles, polystyrene, cigarette_butts and so on. Additionally, there are also variables such as `weight_tons` and `volume_cubic_yards`, which can be used to quantify the total dumpsters collected. 
The total weight of trash collected by Professor Trash Wheel is `r sum(professor_tw_data|>pull(weight_tons))`, and the total number of cigarette butts collected by Gwynnda in July of 2021 is `r gwynnda_tw_data|>filter(year == 2021 & month == "July")|>pull(cigarette_butts)|>sum()|>as.integer()`.



## Problem 3

#### Tidy up Demographic Data

```{r}
mci_demographic_data = read_csv(file = "./data_mci/MCI_Baseline.csv", skip =1)|> #load the demographic csv file
  janitor::clean_names()
```
```{r}
mcidemo_data = mci_demographic_data|>
  janitor::clean_names()|> #recode names to appropriate r style 
  mutate(
    sex = case_match( #this part comes from class note
      sex,
      1 ~ "male",
      0 ~ "female"
    ),
    apoe4 = case_match( #follow the same logic as translating binary numeric values to categorical 
      apoe4,
      1 ~ "carrier",
      0 ~ "non-carrier"
    )
  )|>
  filter(age_at_onset !=".") #remove participants with no MCI at baseline
  
```

#### Discussion

Notably, we translate two numeric binary variable `sex` and `apoe4` to categorical variables:female/male and carrier/non-carrier. In addition, we remove participants whose `age_at_onset` is missing by using the `filter` function. `r nrow(mci_demographic_data)` participants were recruited, and `r nrow(mcidemo_data)` participants developed MCI. The average baseline age is `r mean(pull(mci_demographic_data, current_age))`. The proportion of women who are APOE4 carriers is approximately `r scales::percent(nrow(mci_demographic_data|>filter(sex == 0 & apoe4 == 1)) / nrow(mci_demographic_data|>filter(sex == 0)))` round to the nearest integer.


#### Tidy up Longitudinally Observed Biomarker Values

```{r}
mci_longitude_data = read_csv(file = "./data_mci/mci_amyloid.csv", skip =1)|> #load the demographic csv file
  janitor::clean_names()

amyloid = mci_longitude_data|>
   rename(id = study_id)|>
  pivot_longer(
    baseline:time_8,
    names_to = "visit_time_elapsed",
    values_to = "amyloid_β_ratio"
  )|>
  drop_na(amyloid_β_ratio)
  
```
Similarly, import the csv file and skip line 1 because the first column was the note. Then, rename the variable names to appropriate r style. Also, `study_id` is renamed to `id` in order to merge the two datasets. Additionally, n.a values are omitted after pivoting the data frame longer. 


