---
title: "p8105_hw2_yy3439"
author: "Yuxin Yin"
output: github_document
date: "2023-10-03"
---

```{r}
library(tidyverse)
library(dplyr)
```

## Problem 1

Firstly, import the `pols-month.csv` file

```{r}
pols_data = read_csv(file = "./fivethirtyeight_datasets/pols-month.csv") #import the pols-month csv file

```

#### pols_data



```{r}
pols = pols_data |>
   janitor::clean_names() |> #clean-up names to r suitable names
   separate(mon, into = c("year", "month", "day"), sep = "-") |> #separate the mon variable by "-"
   mutate(
     month = as.numeric(month), #change the char variable month to numeric
     year = as.numeric(year),
     month = month.name[month]) |> #translate numeric month number to full month names
  select(-prez_dem, -prez_gop) |> #remove the prez_dem and prez_gop variables
   pivot_longer(
    gov_gop:rep_dem, #select the remaining columns
     values_to = "president") |> #and assign the values of these variables to a new variable "president"
      select(-day) #remove the day variable
  
```


#### snp_data


```{r}
snp_data = read_csv(file = "./fivethirtyeight_datasets/snp.csv") #load the snp.csv file
```


```{r}
snp = snp_data|>
   janitor::clean_names() |> #clean-up names to r suitable names
  mutate(date = format(as.Date(date, format = "%m/%d/%y"), "%Y/%m/%d"))|>
   separate(date, into = c("year", "month", "day"), sep = "/") |>#separate the date variable into year, month, and day
   mutate(
     year = as.numeric(year),    #change the char variable month to numeric
     year = ifelse(year >2023, year - 100, year),
     month = as.numeric(month),
      month = month.name[month]
    )|>
  select(-day)|>
  arrange(year, month)|>  #rearrange by the order of year and month
  relocate(year, month, everything())
```

#### unemploy_data


```{r}
unemploy_data = read_csv(file = "./fivethirtyeight_datasets/unemployment.csv") #load the snp.csv file
```


```{r}
unemployment = unemploy_data|>
   janitor::clean_names() |>
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemployment"
  )|>
  mutate(
    year = as.numeric(year),
    month = month.name[match(month, tolower(month.abb))]
  )
```
#### Merging Datasets


```{r}
merged_data = left_join(pols, snp, by = c("year", "month")) |>
  left_join(unemployment, by = c("year", "month"))
merged_data
```
The `pols` dataset has `r nrow(pols)` observations after cleaning. This dataset contains data related to president information and party affiliation. In addition, the `snp` dataset has `r nrow(snp)` observations after cleaning, and the `unemployment` dataset has `r nrow(unemployment)` observations after cleaning. The `snp` dataset contains data related to `close` and the `unemployment` dataset contains information related to unemployed rate. The combined dataset has `r nrow(merged_data)` observations and `r ncol(merged_data)` variables after cleaning. Finally, we noticed that parts of both `close` and `unemployment` are missing for the final merged dataset. We can't simply omit all the NAs because some appear in `close` while the others appear in `unemployment`.






## Problem 2


```{r}
library(readxl)
mr_tw_data = read_excel("./fivethirtyeight_datasets/ts.xlsx", sheet = "Mr. Trash Wheel", skip = 1) #load the excel file with Mr. Trash Wheel sheet and omit the first column with figure 
```


#### Tidy Data

```{r}
mr_trash_wheel_data = mr_tw_data |> 
  janitor::clean_names() |> #clean names to appropriate r format
  select(dumpster:homes_powered)|> #omit non-data entries 
  drop_na(dumpster) |> #omit rows with no dumpster-specific data 
  mutate(
    homes_powered = ((weight_tons*500)/30), #add another variable based on the calculation in the notes
    tw_name = "Mr. Trash Wheel", #to keep track of the records after combining datasets 
    year = as.numeric(year) #change the variable type to merge datasets
    
  )

```


Similarly, we follow the same cleanning process to organize the data for Professor Trash Wheel and Gwynnda


```{r}
professor_tw_data = read_excel("./fivethirtyeight_datasets/ts.xlsx", sheet = "Professor Trash Wheel", skip = 1)|>
   janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    homes_powered = ((weight_tons*500)/30),
    tw_name = "Professor Trash Wheel"

  )

gwynnda_tw_data = read_excel("./fivethirtyeight_datasets/ts.xlsx", sheet = "Gwynnda Trash Wheel", skip = 1)|>
   janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    homes_powered = ((weight_tons*500)/30),
     tw_name = "Gwynnda Trash Wheel"
  )
```

#### Combine Datasets

Finally, combining the above three datasets

```{r}
trash_wheel = bind_rows(mr_trash_wheel_data, professor_tw_data, gwynnda_tw_data) |>
  select(trash_wheel_name = "tw_name", everything())
```


#### Discussion

The `Mr Trash Wheel` sheet has `r nrow(mr_trash_wheel_data)` observations. The `Professor Trash Wheel` sheet has `r nrow(professor_tw_data)` observations, and the `Gwynnda Trash Wheel` sheet has `r nrow(gwynnda_tw_data)` observations. Specifically, all these data is calculated after we omit rows with no specific dumpster data. 
In addition, the final combined dataset `trash_wheel` contains `r nrow(trash_wheel)` observations and `r ncol(trash_wheel)` columns. 
Each sheet contains similar key variables specifying various types of dumpsters: plastic bottles, polystyrene, cigarette_butts and so on. Additionally, there are also variables such as `weight_tons` and `volume_cubic_yards`, which can be used to quantify the total dumpsters collected. 
The total weight of trash collected by Professor Trash Wheel is `r sum(professor_tw_data|>pull(weight_tons))`, and the total number of cigarette butts collected by Gwynnda in July of 2021 is `r gwynnda_tw_data|>filter(year == 2021 & month == "July")|>pull(cigarette_butts)|>sum()|>as.integer()`.



## Problem 3

#### Tidy up Demographic Data

```{r}
mci_demographic_data = read_csv(file = "./data_mci/MCI_Baseline.csv", skip =1)|> #load the demographic csv file
  janitor::clean_names()
```
```{r}
baseline_data = mci_demographic_data|>
  janitor::clean_names()|> #recode names to appropriate r style 
  mutate(
    sex = case_match( #this part comes from class note
      sex,
      1 ~ "male",
      0 ~ "female"
    ),
    apoe4 = case_match( #follow the same logic as translating binary numeric values to categorical 
      apoe4,
      1 ~ "carrier",
      0 ~ "non-carrier"
    )
  )|>
  filter(age_at_onset !=".") #remove participants with no MCI at baseline
  
```

#### Discussion

Notably, we translate two numeric binary variable `sex` and `apoe4` to categorical variables:female/male and carrier/non-carrier. In addition, we remove participants whose `age_at_onset` is missing by using the `filter` function. `r nrow(mci_demographic_data)` participants were recruited, and `r nrow(baseline_data)` participants developed MCI. The average baseline age is `r mean(pull(baseline_data, current_age))`. The proportion of women who are APOE4 carriers is approximately `r scales::percent(nrow(baseline_data|>filter(sex == 0 & apoe4 == 1)) / nrow(baseline_data|>filter(sex == 0)))` round to the nearest integer.


#### Tidy up Longitudinally Observed Biomarker Values

```{r}
mci_longitude_data = read_csv(file = "./data_mci/mci_amyloid.csv", skip =1)|> #load the demographic csv file
  janitor::clean_names()|>
  rename(id = study_id)

amyloid = mci_longitude_data|>
  pivot_longer(
    baseline:time_8,
    names_to = "visit_time_elapsed",
    values_to = "amyloid_β_ratio"
  )|>
  drop_na(amyloid_β_ratio)
  
```
Similarly, import the csv file and skip line 1 because the first column was the note. Then, rename the variable names to appropriate r style. Also, `study_id` is renamed to `id` in order to merge the two datasets. Additionally, n.a values are omitted after pivoting the data frame longer. 

Before merging the datasets, we can check participants that only appear in either the baseline or amyloid dataset. We can see that 12 participants appear only in `amyloid` dataset and 8 participants only appear in `baseline` dataset.
```{r}
unique_in_amyloid  = anti_join(mci_longitude_data, mci_demographic_data, by = "id")
unique_in_amyloid

unique_in_baseline  = anti_join(mci_demographic_data, mci_longitude_data, by = "id")
unique_in_baseline
```


#### Merging Datasets


```{r}
full_dataset = inner_join(baseline_data, amyloid, by = c("id"))
```
There are `r nrow(full_dataset)` participants and `r ncol(full_dataset)` columns in the final merged dataset. The average education level is `r mean(full_dataset|>pull(education))`.


#### Exporting

```{r}
write.csv(full_dataset, "./data_mci/mci_full.csv")
```

